Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-08 rigid_bodies.tex
Revision of NVT adjustment

Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined kinetic energy} can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega}$.
In this definition, ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.

@@

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each body $i$, such as required for preserving the unit-norm constraint \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of \cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.

@@

Bringing the attention back to the NVT case, the observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $6N-3$ degrees of freedom is that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T,
\end{equation*}
which becomes $2\langle \refined{K} \rangle = (6N - 3) k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.

@@

One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-2h}$};
\node at (2.2,-0.2)[below]{$r_{t-h}$};
\node at (4.4,-0.2)[below]{$r_{t}$};
\node at (6.6,-0.2)[below]{$r_{t+h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}

In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ goes as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.

Please note that Eq.~\ref{eq:generalized force NVT} is closely related to Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\refined{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\refined{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation*}
By substituting $\refined{\Ham{}}$ in this equation, we obtain that
\begin{equation}
2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation}
This result, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.

Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \refined{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
This is done as follows:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

A drawback of our approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.

>>>>

Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined kinetic energy} can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.

@@

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each body $i$, such as required for preserving the unit-norm constraint \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of \cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each diagonal block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.

@@

Bringing the attention back to the NVT case, the observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = L k_\text{\tiny B} T,
\end{equation*}
which becomes $2\langle \refined{K} \rangle = L k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.

@@

One instant in which $\refined K$ must be evaluated is immediately after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
This takes place once per time step, intercalating with non-Hamiltonian propagations.
Therefore, at such instant there is no previous record of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}$ and $\dot{\vt q}$ with high accuracy.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to generate a four-point stencil by executing two virtual, incomplete Verlet steps, backward and forward in time, such as illustrated in Fig.~\ref{fig:virtual steps}.
By virtual and incomplete we mean, respectively, that these steps are never incorporated to the trajectory and that they are aborted once the positions and orientations have been updated.
By this means, we can obtain estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.

\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{${\vt r}(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{${\vt r}_{t-2h}$};
\node at (2.2,-0.2)[below]{${\vt r}_{t-h}$};
\node at (4.4,-0.2)[below]{${\vt r}_{t}$};
\node at (6.6,-0.2)[below]{${\vt r}_{t+h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual steps}
\end{figure}

In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales both $\vt p$ and $\vt \pi$ by a factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
Because such action leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ is done as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.

Finally, given that configurations $\vt x$ are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}(\vt x)}$, we need to reweight any computed observable $A(\vt x)$ so that its average is consistent with the original distribution.
Such reweighting is done by \cite{Torrie_1977}
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

A drawback of our approach is that the resulting integrator lacks time-reversal symmetry due to the asymmetric differentiation formulas.
As a matter of fact, long-term reversibility is not generally achievable with software that relies on floating-point arithmetics.
In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per time step.
Higher-order derivative estimates can also be obtained in this fashion if necessary.
